---
title: "Plot coverage around annotated starts"
author: "`r Sys.info()[['user']]`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    toc: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import libraries
```{r, message=F}
# genome sequence
library(BSgenome.Hsapiens.UCSC.hg38)
# for working with annotations
library(GenomicFeatures)
# for working with bam files
library(GenomicAlignments)
# read/write gff3 files
library(rtracklayer)
# for tab-data manipulations
library(tidyverse)
# for easy string concatentation
library(glue)
# for converting bioconductor objects to tidyverse
library(biobroom)
# for using tidyverse on bioconductor objects
library(plyranges)
```

## Analysis specific parameters
```{r}
sample_names <- c(
  "vir" = "+vir",
  "ifn_vir" = "+ifn +vir",
  "untr" = "ctrl",
  "ifn" = "+ifn"
)

treatment_names <- c(
  "cyclo" = "ribo",
  "ltm" = "ribo + LTM",
  "mrna" = "mRNA"
)

min.read.density <- 0.33  # reads / nt, only cds above this used for codon density
left_overhang_for_start_profile <- 50
right_overhang_for_start_profile <- 100

left_overhang_for_stop_profile <- 100
right_overhang_for_stop_profile <- 50

```

## Plotting defaults
```{r, message=F}
# color blind palette
# http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
cbPalette <- c("#666666", "#E69F00", "#56B4E9", "#009E73", "#CC79A7", "#0072B2", 
    "#D55E00", "#F0E442")

theme_set(theme_classic(base_family = "Helvetica", base_size = 8) +
theme(          
  strip.background = element_blank(),
  legend.text = element_text(size = 8),
  strip.text.x = element_text(size = 8),
  axis.line = element_line(size=.25),
  plot.title = element_text(size = 8, hjust = 0.5)
))
```

## Read in genome and annotations
```{r}
genome <- BSgenome.Hsapiens.UCSC.hg38
canonical <- import.gff3(glue(
  '/fh/fast/subramaniam_a/db/rasi/genomes/human/hg38/gencode/',
  'annotations/',
  'gencode.v24.canonical_ccds_transcripts.20170315.gff3.gz'))

tx <- canonical %>% 
  filter(type == "exon") %>% 
  split(.$transcript_id) 

cds <- canonical %>% 
  filter(type == "CDS") %>% 
  split(.$transcript_id) 

tx_annotations <- tx %>% 
  unlist() %>% 
  tidy() %>% 
  group_by(transcript_id) %>% 
  summarize(tx_length = sum(width)) 

cds_annotations <- canonical[canonical$type == "CDS"] %>% 
  tidy() %>% 
  group_by(transcript_id) %>% 
  summarize(cds_length = sum(width), gene_id = first(gene_id), gene_name = first(gene_name)) %>% 
  left_join(tx_annotations, by = "transcript_id") %>% 
  print()
```


## Get list of all annotated starts
```{r, warning=F, message=F, echo=F}
starts <- canonical %>% 
  filter(type == "start_codon") %>% 
  # map to transcriptomic coordinates
  GenomicFeatures::mapToTranscripts(tx) %>% 
  # set strand to be positive since we are in transcript coords
  mutate(strand = "+")


starts <- starts %>% 
  # extend to either side for plotting profile
  # note that regions outside of transcript boundaries will be automatically
  # discarded at the inner join with the read density below
  promoters(upstream = left_overhang_for_start_profile, 
            downstream = right_overhang_for_start_profile) %>% 
  tidy() %>%
  # create a sequence from start to stop for each range
  mutate(pos = map2(start, end, function(x, y) seq(from = x, to = y))) %>%
  # expand each range to equal its length
  unnest()  %>%
  mutate(loc = pos - start) %>%
  # mutate and unnest to create a single pos for each location
  mutate(start = pos, end = pos) %>%
  select(start, end, seqname, loc) %>%
  left_join(select(cds_annotations, transcript_id, gene_name), by = c("seqname" = "transcript_id")) %>% 
  print()
```

## Get canonical start tx coord for identifying frame below
```{r}
starts_tx <- canonical %>% 
  filter(type == "start_codon") %>% 
  tidy() %>% 
  mutate(end = if_else(strand == "+", start, end)) %>%
  mutate(start = if_else(strand == "-", end, start)) %>%
  select(seqname, start, end, strand) %>% 
  GRanges() %>% 
  mapToTranscripts(tx) %>%
  tidy() %>% 
  group_by(xHits) %>% 
  arrange(start) %>% 
  slice(1) %>% 
  ungroup() %>% 
  mutate(cds.start = start) %>% 
  select(cds.start, seqname) %>% 
  print()
```


## Get list of all annotated stops
```{r, results="hide", warning=F}
stops <- canonical %>% 
  filter(type == "stop_codon") %>% 
  # map to transcriptomic coordinates
  GenomicFeatures::mapToTranscripts(tx)

# set strand to be + since we are now in tx space
strand(stops) <- "+"

stops <- stops %>% 
  # extend to either side for plotting profile
  # note that regions outside of transcript boundaries will be automatically
  # discarded at the inner join with the read density below
  promoters(upstream = left_overhang_for_stop_profile, 
            downstream = right_overhang_for_stop_profile) %>% 
  tidy() %>% 
  # create a sequence from start to stop for each range
  mutate(pos = map2(start, end, function(x, y) seq(from = x, to = y))) %>% 
  # expand each range to equal its length
  unnest()  %>% 
  mutate(loc = pos - start) %>% 
  # mutate and unnest to create a single pos for each location
  mutate(start = pos, end = pos) %>% 
  select(start, end, seqname, loc) %>% 
  print()
```

## Read in coverage
```{r}
cvg <- list.files("../coverage", full.names = T, recursive = T) %>% 
  # get covearge file
  str_subset("gencode.+tsv.gz$") %>% 
  # extract samplename from filename
  setNames(str_extract(., "[^/\\.]+(?=\\.gencode)")) %>% 
  enframe("sample", "file") %>%
  # exclude deblurred or pooled samples
  filter(!str_detect(file, "deblurred|pooled")) %>% 
  mutate(data = map(file, function(x) suppressMessages(read_tsv(x)))) %>%
  select(-file) %>%
  unnest() %>%
  mutate(sample = as.factor(sample)) %>%
  GRanges() %>%
  print()
```

```{r}
cds_counts <- mapToTranscripts(cvg, cds) %>% 
  mutate(score = cvg$score[xHits], sample = cvg$sample[xHits]) %>% 
  tidy() %>% 
  group_by(sample, seqname) %>%
  summarize(cds_counts = sum(score * width)) %>% 
  ungroup() %>%
  left_join(select(cds_annotations, transcript_id, cds_length),
            by = c("seqname" = "transcript_id")) %>% 
  mutate(avg_read_density = cds_counts / cds_length) %>% 
  # select(sample, seqname, avg_read_density) %>% 
  arrange(desc(avg_read_density)) %>% 
  print()

```

## Convert coverage to transcriptome coordinates and normalize by mean ORF density
```{r}
# go to tx coords
cvg_tx <- mapToTranscripts(cvg, tx)
cvg_tx$score  <- cvg$score[cvg_tx$xHits]
cvg_tx$sample <- cvg$sample[cvg_tx$xHits]

### Thresolding and normalization
# normalize each read by the average read density / nt in the cds region of each transcript
normalized_density <- cvg_tx %>% 
  tidy() %>% 
  filter(score > 0) %>% 
  left_join(cds_counts, by = c("seqname", "sample")) %>% 
  mutate(norm_density = score / avg_read_density)  %>% 
  filter(avg_read_density > min.read.density) %>% 
  # convert each range of length > 1 to a sequence of ranges of length 1
  mutate(pos = map2(start, end, function(x, y) seq(x, y)))  %>% 
  unnest() %>% 
  mutate(start = pos, end = pos) %>% 
  select(-pos) %>% 
  print()
```


## Calculate read density in each frame
```{r}
frame_density <- normalized_density %>% 
  inner_join(starts_tx, by = "seqname") %>% 
  mutate(frame = (start - cds.start) %% 3) %>% 
  # divide by cds length to account for zero count positions
  mutate(frame_density = norm_density / cds_length) %>% 
  group_by(sample, frame) %>% 
  # the sum here takes care of weighin down the zero count positions
  summarize(frame_density = sum(frame_density)) %>% 
  ungroup() %>% 
  # normalize
  group_by(sample) %>% 
  mutate(frame_density = frame_density / sum(frame_density) * 100) %>% 
  ungroup() %>% 
  mutate(treatment = str_extract(sample, "mrna|cyclo|ltm")) %>% 
  mutate(sample = str_extract(sample, "(?<=(mrna|cyclo|ltm)_).+")) %>% 
  mutate(sample = forcats::fct_rev(sample_names[sample]), 
         treatment = treatment_names[treatment]) %>% 
  write_tsv(glue('../tables/frame_density_host.tsv.gz'))
  
```

## Plot density in each frame
```{r, fig.width = 3.6, fig.height = 1.8}
frame_density %>% 
  ggplot(aes(x = frame, y = frame_density, color = treatment)) +
  facet_wrap(~sample, ncol = 4, scales = "free") +
  geom_line(size = 0.5) + geom_point() +
  labs(x = "frame", 
       y = "normalized read density (%)",
       color = "") +
  scale_color_manual(values = cbPalette) +
  scale_x_continuous(breaks = c(0,1,2)) +
  theme(legend.position = "top", legend.direction = "horizontal")

ggsave(glue('../figures/frame_density_profile_host.pdf'))

```


## Plot profile around start codon
### Calculate profile
```{r}
plot_data <- normalized_density %>% 
  right_join(starts, by = c("seqname", "start", "end")) %>% 
  # gets rid of positions outside tx boundary
  filter(start > 0) %>%
  select(sample, seqname, loc, norm_density) %>%
  complete(sample, nesting(seqname, loc), fill = list(norm_density = 0)) %>%
  # at least one position in the window must be greater than 0 in a given sample
  # (this gets rid of NA sample created in the right_join above)
  group_by(sample, seqname) %>%
  filter(sum(norm_density) > 0) %>%
  ungroup() %>%
  # adjust location to account for overhang
  mutate(loc = loc - left_overhang_for_start_profile) %>% 
  group_by(sample, loc) %>% 
  # average each location across all genes in each sample
  summarize(norm_density = mean(norm_density)) %>%
  ungroup() %>% 
  mutate(treatment = str_extract(sample, "mrna|cyclo|ltm")) %>% 
  mutate(sample = str_extract(sample, "(?<=(mrna|cyclo|ltm)_).+")) %>% 
  mutate(sample = forcats::fct_rev(sample_names[sample]), 
         treatment = treatment_names[treatment]) %>% 
  write_tsv(glue('../tables/start_codon_profile_host.tsv.gz'))
```

### Plot profile
```{r, fig.width=2.85, fig.height=5.6}
plot_data %>% 
  filter(loc %in% seq(-20, 30)) %>%
  ggplot(aes(x = loc, y = norm_density, color = treatment)) +
  facet_wrap(sample ~ treatment, ncol = 3, scales = "free") +
  geom_line(size = 0.5) + 
  labs(x = "distance from start codon (nt)", 
       y = "normalized read density",
       color = "") +
  scale_color_manual(values = cbPalette) +
  scale_x_continuous(breaks = c(-15, 0, 15, 30)) +
  scale_y_continuous(limits = c(0, NA)) +
  theme(legend.position = "top", legend.direction = "horizontal")

ggsave(glue('../figures/start_codon_profile_host.pdf'))
```

### Same plot as above but expanded to show 3nt periodicity
```{r, fig.width=2.85, fig.height=5.6}
plot_data %>% 
  filter(loc %in% seq(50, 80)) %>%
  ggplot(aes(x = loc, y = norm_density, color = treatment)) +
  facet_wrap(sample ~ treatment, ncol = 3, scales = "free") +
  geom_line(size = 0.5) + 
  labs(x = "distance from start codon (nt)", 
       y = "normalized read density",
       color = "") +
  scale_color_manual(values = cbPalette) +
  # scale_x_continuous(breaks = c(-15, 0, 15, 30)) +
  scale_y_continuous(limits = c(0, NA)) +
  theme(legend.position = "top", legend.direction = "horizontal")

ggsave(glue('../figures/start_codon_profile_host_3nt_periodicity.pdf'))
```
## Plot profile around stop codon 
### Calculate profile
```{r}
plot_data <- normalized_density %>% 
  right_join(stops, by = c("seqname", "start", "end")) %>% 
  # gets rid of positions outside tx boundary
  filter(start > 0) %>%
  select(sample, seqname, loc, norm_density) %>%
  complete(sample, nesting(seqname, loc), fill = list(norm_density = 0)) %>%
  # at least one position in the window must be greater than 0 in a given sample
  # (this gets rid of NA sample created in the right_join above)
  group_by(sample, seqname) %>%
  filter(sum(norm_density) > 0) %>%
  ungroup() %>%
  # adjust location to account for overhang
  mutate(loc = loc - left_overhang_for_stop_profile) %>% 
  group_by(sample, loc) %>% 
  # average each location across all genes in each sample
  summarize(norm_density = mean(norm_density)) %>%
  ungroup() %>% 
  mutate(treatment = str_extract(sample, "mrna|cyclo|ltm")) %>% 
  mutate(sample = str_extract(sample, "(?<=(mrna|cyclo|ltm)_).+")) %>% 
  mutate(sample = forcats::fct_rev(sample_names[sample]), 
         treatment = treatment_names[treatment]) %>% 
  write_tsv(glue('../tables/stop_codon_profile_host.tsv.gz'))
```

### Plot profile
```{r, fig.width=2.85, fig.height=5.6}
plot_data %>% 
  filter(loc %in% seq(-30, 20)) %>%
  ggplot(aes(x = loc, y = norm_density, color = treatment)) +
  facet_wrap(sample ~ treatment, ncol = 3, scales = "free") +
  geom_line(size = 0.5) + 
  labs(x = "distance from stop codon (nt)", 
       y = "normalized read density",
       color = "") +
  scale_color_manual(values = cbPalette) +
  scale_x_continuous(breaks = c(-30, -15, 0, 15)) +
  scale_y_continuous(limits = c(0, NA)) +
  theme(legend.position = "top", legend.direction = "horizontal")

ggsave(glue('../figures/stop_codon_profile_host.pdf'))
```
